Computing the {\em least-commitment} action model from state observations

### 2018-11-29 doc/ijcai-19-conformant/conformant-learning.tex
small updates added to ijcai19 paper
added definition for partially specified model

\subsection{The {\em least-commitment} action model}
The task of computing the {\em least-commitment} action model from state observations is defined as a tuple $\tup{P,M,\mathcal{O}}$ where:
\begin{itemize}
\item $P=\tup{F,A[\cdot],I,G}$ is a planning problem where $A[\cdot]$ is a set of actions s.t. for each $a\in A[\cdot]$, the semantics of $a$ is unknown; i.e. the functions $\rho$ and/or $\theta$ of $a$ are undefined. 
\item $M$ is the set of possible models for the actions in $A[\cdot]$. A model $\mathcal{M}\in M$ defines the semantics of every action in $A[\cdot]$. Planning models $\mathcal{M}\in M$ differ in the $\tup{\rho,\theta}$ functions of the actions but they all use the same set of state variables $F$.
\item $\mathcal{O}(\tau)$ is a sequence of partial states coming from the observation of a trajectory $\tau(\pi,s_0)$ produced by the execution of certain unknown plan $\pi$ that solves $P$.
\end{itemize}

The {\em solution} to this task is the {\em least-commitment} action model, before formalizing it we introduce the following necessary definition.

\begin{definition}[Explanation of an observation]
Given a model $\mathcal{M}\in M$ for the actions $A[\cdot]$ in $P$, and a sequence of partially observed states $\mathcal{O}(\tau)$, we say that {\em the model explains the observation} (denoted by $\mathcal{M}\mapsto\mathcal{O}(\tau)$) iff there exists a plan $\pi$ that solves $P$ and that it is consistent with $\pi$. We say that $\pi$ is the {\em best explanation} for $\mathcal{O}(\tau)$ iff, in additon, $\pi$ is optimal.
\end{definition}

Now we are ready to define the {\em least-commitment} action model for an observation $\mathcal{O}(\tau)$ as the largest subset of models $M^*\subseteq M$ such that every model $\mathcal{M}\in M^*$ {\em explains} the input observation.

@@

\subsection{Computing the {\em least-commitment} model via conformant planning}
Given a sequence of partial states $\mathcal{O}(\tau)$, and a set of actions $A[\cdot]$ with known list of parameters, but unknown $\rho$ and $\theta$ functions, we can build a classical planning problem $P=\tup{F,A[\cdot],I,G}$ such that $\mathcal{O}(\tau)$ represents the observation of a $\tau(\pi,I)$ trajectory that solves $P$. In this section we show that starting from $\mathcal{O}(\tau)$ and $P$ we can build a {\em conformant planning problem} $P_c$ whose solution induces the {\em least-commitment} action model for observation $\mathcal{O}(\tau)$.

In more detail, we build a {\em conformant planning problem} $P_c=\tup{F_c,A_c,\Upsilon,G}$ such that:
\begin{itemize}
\item The set of fluents $F_c$ extends $F$ with two new sets of fluents:
      \begin{itemize}
      \item $\{test_j\}_{1\leq j\leq m}$, indicating the state observation $s_j\in\mathcal{O}(\tau)$ where the action model is validated
      \item The fluents {\tt\small pre\_e\_$\xi$} and {\tt\small eff\_e\_$\xi$} (where $e\in{\mathcal I}_{\Psi,\xi}$) for a propositional encoding of the {\em preconditions}, {\em negative}, and {\em positive} effects of an action schema $\xi$. Our encoding exploits the syntactic constraint of \strips\ so, if {\tt\small pre\_e\_$\xi$} and {\tt\small eff\_e\_$\xi$} holds it means that $e\in{\mathcal I}_{\Psi,\xi}$ is a negative effect in $\xi$ and if $pre\_e\_\xi$ does not hold but {\tt\small eff\_e\_$\xi$} holds, it means that $e\in{\mathcal I}_{\Psi,\xi}$ is a positive effect in $\xi$. Figure~\ref{fig:propositional} shows the PDDL encoding of the {\tt\small stack(?v1,?v2)} schema and our propositional representation for this same schema.
      \end {itemize}

@@

\item The clauses in $\Upsilon$ extend the set of fluents that hold in the initial state $I=s_0$ and $mode_{prog}$ set to true. In addition, $\Upsilon$ includes the two clauses {\tt\small pre\_e\_$\xi$} xor {\tt\small $\neg$pre\_e\_$\xi$} and {\tt\small eff\_e\_$\xi$} xor {\tt\small $\neg$eff\_e\_$\xi$} for every $\xi$ and $e\in{\mathcal I}_{\Psi,\xi}$. This means that the actual value of these fluents is unknown, so any model from the \strips\ space of models can initially be part of the {\em least-commitment} action model.

@@

\section{Conclusions}

>>>>

\subsection{The {\em least-commitment} action model}
The task of computing the {\em least-commitment} action model from state observations is defined as a $\tup{P,\mathcal{O}}$ pair:
\begin{itemize}
\item  $P=\tup{F,A[\cdot],I,G}$ is a planning problem where $A[\cdot]$ is a set of actions s.t. for each $a\in A[\cdot]$, the semantics of $a$ is unknown; i.e. the corresponding $\tup{\rho,\theta}$ functions are undefined. 
\item $\mathcal{O}(\tau)$ is a sequence of partial states coming from the observation of a trajectory $\tau(\pi,s_0)$ produced by the execution of certain unknown plan $\pi$ that solves $P$.
\end{itemize}

Before formalizing the solution to this task, i.e. the {\em least-commitment} action model, we introduce several necessary definitions. We first start defining a {\em partially specified action model} inspired by the notion of {\em incomplete (annotated) model}~\cite{sreedharan2018handling}.
\begin{definition}[Partially specified action model]
Given a set of actions $A[\cdot]$ and a set of fluents $F$ then, a {\em partially specified action model} $M$ is a set of possible models for the actions in $A[\cdot]$ such that: (1), any model $\mathcal{M}\in M$ defines the $\tup{\rho,\theta}$ functions of every action in $A[\cdot]$ and (2), for every $\mathcal{M}\in M$ the $\tup{\rho,\theta}$ functions are defined in the set of state variables $F$. (Note that if $M$ is a singleton it represents a {\em fully specified action model}).
\end{definition}

\begin{definition}[Explanation of an observation]
Given a fully specified action model $\mathcal{M}$ for the actions $A[\cdot]$ in $P$, and a sequence of partially observed states $\mathcal{O}(\tau)$, we say that {\em the model explains the observation} (denoted $\mathcal{M}\mapsto\mathcal{O}(\tau)$) iff there exists a plan $\pi$, consistent with $\mathcal{O}(\tau)$, that solves $P$. We say that $\pi$ is the {\em best explanation} for $\mathcal{O}(\tau)$ iff, in addition, $\pi$ is optimal. 
\end{definition}

Now we are ready to define the {\em least-commitment} action model for an observation $\mathcal{O}(\tau)$.
\begin{definition}[The {\em least-commitment} action model]
Given a {\em partially specified action model} $M$ that represents the space of possible action models and a sequence of partially observed states $\mathcal{O}(\tau)$, then the {\em least-commitment} action model is another {\em partially specified action model} that represents the largest subset of models $M^*\subseteq M$ such that every model $\mathcal{M}\in M^*$ {\em explains} the input observation.
\end{definition}

@@

\subsection{Computing the {\em least-commitment} model via conformant planning}
Given the set of actions $A[\cdot]$ with unknown $\rho$ and $\theta$ functions and a sequence of partial states $\mathcal{O}(\tau)$, we can build a classical planning problem $P=\tup{F,A[\cdot],I,G}$ such that $\mathcal{O}(\tau)$ represents the observation of a $\tau(\pi,I)$ trajectory that solves $P$.

In this section we show that starting from $\mathcal{O}(\tau)$ and $P$ we can build a {\em conformant planning problem} $P_c$ whose solution induces the {\em least-commitment} action model for observation $\mathcal{O}(\tau)$. In more detail, we build a {\em conformant planning problem} $P_c=\tup{F_c,A_c,\Upsilon,G}$ such that:
\begin{itemize}
\item The set of fluents $F_c$ extends $F$ with two new sets of fluents:
      \begin{itemize}
      \item $\{test_j\}_{1\leq j\leq m}$, indicating the state observation $s_j\in\mathcal{O}(\tau)$ where the action model is validated
      \item The fluents {\tt\small pre\_e\_$\xi$} and {\tt\small eff\_e\_$\xi$} (where $e\in{\mathcal I}_{\Psi,\xi}$) for a propositional encoding of the {\em preconditions}, {\em negative}, and {\em positive} effects of an action schema $\xi$. Our encoding exploits the syntactic constraint of \strips\ so, if {\tt\small pre\_e\_$\xi$} and {\tt\small eff\_e\_$\xi$} holds it means that $e\in{\mathcal I}_{\Psi,\xi}$ is a negative effect in $\xi$ and if $pre\_e\_\xi$ does not hold but {\tt\small eff\_e\_$\xi$} holds, it means that $e\in{\mathcal I}_{\Psi,\xi}$ is a positive effect in $\xi$. Figure~\ref{fig:propositional} shows the PDDL encoding of the {\tt\small stack(?v1,?v2)} schema and our propositional representation for this same schema.
      \end {itemize}

@@

\item The clauses in $\Upsilon$ extend the set of fluents that hold in the initial state $I=s_0$ and $mode_{commit}$ set to true. In addition, $\Upsilon$ includes the two clauses {\tt\small pre\_e\_$\xi$} xor {\tt\small $\neg$pre\_e\_$\xi$} and {\tt\small eff\_e\_$\xi$} xor {\tt\small $\neg$eff\_e\_$\xi$} for every $\xi$ and $e\in{\mathcal I}_{\Psi,\xi}$. This means that the actual value of these fluents is unknown, so any model from the \strips\ space of models can initially be part of the {\em least-commitment} action model.

@@

\section{Conclusions}
~\cite{SternJ17}
